### 2024-05-16 thesis/input_files/metrics.tex
May 15th 2024

Proper evaluation methods guide researchers in choosing the model that best fits their needs; thus, this chapter is dedicated to the most common evaluation metrics adopted by academics in the field of electricity forecasting. Error metrics and measures vary depending on whethter we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{Mean absolute error}\label{mae}
Consider the time series with actual values given by $y=(y_{n+1}, y_{n+2},\dots, y_{n+h})$
and its h step ahead point forecast $\hat{y}=(\hat{y}_{n+1}, \hat{y}_{n+2},\dots, \hat{y}_{n+h})$ the mean absolute error (MAE) is defined as
\begin{definition}
    $\mathrm{MAE}(y,\hat{y})=\frac{1}{h}\| y- \hat{y}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|y_{n+k}-\hat{y}_{n+k}|$
\end{definition}

\section{Root mean squared error}\label{rmse}
\begin{definition}
    $\mathrm{RMSE(y, \hat{y})}=\frac{1}{\sqrt{h}}\|y-\hat{y}\|_{2}=\sqrt{\frac{\sum\limits_{k=1}^{h}(y_{n+k}- \hat{y}_{n+k})^2}{h}}$
\end{definition}

@@

\section{Root mean squared percentage error}\label{rmspe}
\begin{definition}
    $\mathrm{RMSPE}(y,\hat{y})=100\cdot\sqrt{\frac{1}{h}\sum\limits_{k=1}^{h} \left(\frac{|y_{n+k}-\hat{y}_{n+k}|}{|y_{n+k}|}\right)^2}$
\end{definition}
Note, mean absolute percentage error and root mean squared percentage error may not be appropriate for series which have zero or very small values, for example, electricity demand at the household level; the result is a large score regardless of the absolute errors.
Scaled errors constitute a robust family of scores.
\section{Mean absolute scaled error}\label{mase}
\begin{definition}
    $\mathrm{MASE}(y,\hat{L})=\frac{1}{h}\sum\limits_{k=1}^N\frac{|y_{n+k}-\hat{y}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|y_{k}-y_{k-1}|}$
\end{definition}
In the denominator we have the error of the naïve/persistence model. 
In this model, the current demand makes up the prediction for the next time step; that is $\hat{y}^{\mathrm{naive}}_{n+1}=y_{n}$.
\section{Root mean squared scaled error}\label{rmsse}
\begin{definition}
    $\mathrm{RMSSE}(y,\hat{y})=\sqrt{\frac{1}{h}\sum\limits_{k=1}^N\left(\frac{|y_{n+k}-\hat{y}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|y_{k}-y_{k-1}|}\right)^2}$
\end{definition}

@@

(q-1)(y_{t}-\hat{y}_{t,q}) & y_t > \hat{y}_{t,q} \\

@@

Where delta is the prediction interval (PI) width, that is $\delta=b_t-a_t$ where $u$ is the prediction interval upper threshold and $l$ is the prediction interval lower threshold. This score penalizes observations falling outside the prediction interval and rewards narrow prediction intervals.

@@

    $\mathrm{CRPS}(y, \hat{F})=\int\limits_{-\infty}^{\infty}\left(\hat{F}(x)-\mathbb{1}(x-y) \right)^2 dx$

@@

    $\mathbb{1}(z)=

@@

Notice that, in our case, equation \ref{def_crps}, the distribution $G$ of $Y_1$ and $Y_2$ is degenerate, with all probability mass on a single point.

@@

    \int\limits_{-\infty}^{\infty}\left(\hat{F}(x)-\mathbb{1}(x-L) \right)^2 dx=\mathbb{E}(|X_1-y|)-\frac{1}{2}\mathbb{E}(|X_1-X_2|)

@@

The probability integral transform (PIT) is a method to assess visually the quality of probabilistic a forecasts. PIT is obtained by applying the predicted cumulative density function $\hat{F}$ to your data; if applying such CDF to the data results in a uniform distributed PIT, then $\hat{F}$ is a valid prediction. If not, $\hat{F}$ is not the suited CDF for the considered data. Figure \ref{fig:pit} provides an example, applying the true CDF results in a well calibrated PIT (left). Alternatively, applying a bad CDF results in either a overdispersed (middle) or underdispersed (right) PIT.

@@

    \caption{PIT types \cite{haben2023core}}

>>>>

Proper evaluation methods guide researchers in choosing the model that best fits their needs; thus, this chapter is dedicated to the most common evaluation metrics adopted by academics in the field of electricity forecasting. Error metrics and measures vary depending on whethter we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{Mean absolute error}\label{mae}
Consider the time series with actual values given by $y=(y_{n+1}, y_{n+2},\dots, y_{n+h})$
and its $h$ step ahead point forecast $\hat{y}=(\hat{y}_{n+1}, \hat{y}_{n+2},\dots, \hat{y}_{n+h})$ the mean absolute error (MAE) is defined as
\begin{definition}
    $\mathrm{MAE}(y,\hat{y})=\frac{1}{h}\| y- \hat{y}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|y_{n+k}-\hat{y}_{n+k}|$
\end{definition}

\section{Root mean squared error}\label{rmse}
\begin{definition}
    $\mathrm{RMSE}(y, \hat{y})=\frac{1}{\sqrt{h}}\|y-\hat{y}\|_{2}=\sqrt{\frac{\sum\limits_{k=1}^{h}(y_{n+k}- \hat{y}_{n+k})^2}{h}}$
\end{definition}

@@

\section{Root mean squared percentage error}\label{rmspe}
\begin{definition}
    $\mathrm{RMSPE}(y,\hat{y})=100\cdot\sqrt{\frac{1}{h}\sum\limits_{k=1}^{h} \left(\frac{|y_{n+k}-\hat{y}_{n+k}|}{|y_{n+k}|}\right)^2}$
\end{definition}
Note, mean absolute percentage error and root mean squared percentage error may not be appropriate for series which have zero or very small values, for example, electricity demand at the household level; the result is a large score regardless of the absolute errors.
Scaled errors constitute a robust family of scores.
\section{Mean absolute scaled error}\label{mase}
\begin{definition}
    $\mathrm{MASE}(y,\hat{y})=\frac{1}{h}\sum\limits_{k=1}^N\frac{|y_{n+k}-\hat{y}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|y_{k}-y_{k-1}|}$
\end{definition}
In the denominator we have the error of the naïve/persistence model. 
In this model, the current demand makes up the prediction for the next time step; that is $\hat{y}^{\mathrm{naive}}_{n+1}=y_{n}$.
\section{Root mean squared scaled error}\label{rmsse}
\begin{definition}
    $\mathrm{RMSSE}(y,\hat{y})=\sqrt{\frac{1}{h}\sum\limits_{k=1}^N\left(\frac{|y_{n+k}-\hat{y}_{n+k}|}{\frac{1}{h-1}\sum\limits_{k=2}^{h}|y_{k}-y_{k-1}|}\right)^2}$
\end{definition}

@@

(q-1)(\hat{y}_{t,q}-y_{t}) & y_t > \hat{y}_{t,q} \\

@@

Where delta is the prediction interval (PI) width, that is $\delta=u_t-l_t$ where $u_t$ is the prediction interval upper threshold and $l_t$ is the prediction interval lower threshold at time step $t$. This score penalizes observations falling outside the prediction interval and rewards narrow prediction intervals.

@@

    $\mathrm{CRPS}(y, \hat{F})=\int\limits_{-\infty}^{\infty}\left(\hat{F}(x)-\mathbb{I}_{\{x-y\}} \right)^2 dx$

@@

    $\mathbb{I}_{\{z\}}=

@@

Notice that, in our case, equation \ref{def_crps}, the distribution $G$ of $Y_1$ and $Y_2$ is degenerate, with all probability mass on a single point $y$.

@@

    \int\limits_{-\infty}^{\infty}\left(\hat{F}(x)-\mathbb{I}_{\{x-y\}} \right)^2 dx=\mathbb{E}(|X_1-y|)-\frac{1}{2}\mathbb{E}(|X_1-X_2|)

@@

The probability integral transform (PIT) is a method to assess visually the quality of a probabilistic forecast. PIT is obtained by applying the predicted cumulative density function $\hat{F}$ to your data; if applying such CDF to the data results in a uniform distributed PIT, then $\hat{F}$ is a valid prediction. If not, $\hat{F}$ is not the suited CDF for the considered data. Figure \ref{fig:pit} provides an example, applying the true CDF results in a well calibrated PIT (left). Alternatively, applying a bad CDF results in either a overdispersed (middle) or underdispersed (right) PIT.

@@

    \caption{Probability integral transform types \cite{haben2023core}}
